Tight Generalization Bounds for Hard Parameter Sharing in High-dimensional Linear Regression

### 2020-10-14 arxiv/conclusions.tex
various polish

This work proposed a formal study of a widely used hard paramaeter sharing approach in high-dimensional linear regression, where the sample size and feature dimension increase at a fixed ratio.
We provided tight generalization bounds that scale with dataset properties such as the sample sizes and covariate shifts.
Based on these bounds, we analyzed the impact of varying sample sizes and covariate shifts on the prediction loss of hard parameter sharing.
Our work rigorously explains several empirical phenomena such as sample efficiency and negative transfer related to these dataset properties.
We validated our results and conducted further studies on a real world classification task.

We describe several open questions for future work.
First, our bound in Corollary \ref{cor_MTL_loss} involves an error term that scales down with $n_1$.
Tightening this error bound requires showing the limit of $\normFro{({Z^{(1)}}^{\top} Z^{(1)} + {Z^{(2)}}^{\top} Z^{(2)})^{-1} {Z^{(1)}}^{\top} Z^{(1)}}^2$ for two isotropic sample covariance matrices.
This requires studying the asymptotic singular values distribution of the non-symmetric matrix $({Z^{(1)}}^{\top} Z^{(1)})^{-1}{Z^{(2)}}^{\top} Z^{(2)}+\id$, which is still an open problem in random matrix theory.  
%FY: $+\id$ is very important and makes the problem very hard; otherwise the problem can be solved with current RMT methods.
The eigenvalue distribution of this matrix, which has been obtained in \citet{Fmatrix}, might be helpful towards resolving this problem.
%but its singular values will follow a different distribution since the matrix is not symmetric.
 %might require new techniques beyond the current ones in random matrix theory .%\HZ{to add}.
Second, it would be interesting to extend our results to classification problems.
Several recent work have made remarkable progress for logistic regression in the high-dimensional  setting, e.g. \citet{sur2019modern}.
An interesting question is to study logistic regression in a multiple-sample setting.

>>>>

This work proposed a formal study of a widely used hard paramaeter sharing approach in multi-task learning.
We provided sharp bias-variance tradeoffs of HPS given multiple linear gression tasks.
Based on these results, we analyzed how varying sample sizes and covariate shifts impacts HPS.
Our work rigorously explains several empirical phenomena such as negative transfer and covariate shift related to these dataset properties.
We validated our theory and conducted further studies on a real world classification task.
We describe several open questions for future work.
First, it would be interesting to tighten our result in Corollary \ref{cor_MTL_loss}, which would extend the observation in Figure \ref{fig_size} to small $n_1$.
Second, it would be very interesting to extend our result to classification problems such as logistic regression.
%First, our bound in Corollary \ref{cor_MTL_loss} involves an error term that scales down with $n_1$.
%Tightening this error bound requires showing the limit of $\normFro{({Z^{(1)}}^{\top} Z^{(1)} + {Z^{(2)}}^{\top} Z^{(2)})^{-1} {Z^{(1)}}^{\top} Z^{(1)}}^2$ for two isotropic sample covariance matrices.
%This requires studying the asymptotic singular values distribution of the non-symmetric matrix $({Z^{(1)}}^{\top} Z^{(1)})^{-1}{Z^{(2)}}^{\top} Z^{(2)}+\id$, which is still an open problem in random matrix theory.
%FY: $+\id$ is very important and makes the problem very hard; otherwise the problem can be solved with current RMT methods.
%The eigenvalue distribution of this matrix, which has been obtained in \citet{Fmatrix}, might be helpful towards resolving this problem.
%but its singular values will follow a different distribution since the matrix is not symmetric.
 %might require new techniques beyond the current ones in random matrix theory .%\HZ{to add}.
%Second, it would be interesting to extend our results to classification problems.
%Several recent work have made remarkable progress for logistic regression in the high-dimensional  setting, e.g. \citet{sur2019modern}.
%An interesting question is to study logistic regression in a multiple-sample setting.

### 2020-10-14 arxiv/diff_cov.tex
typos
various polish

\begin{figure*}[!t]
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{figures/same_covariates.eps}
		\caption{Example \ref{ex_same_cov}}
		\label{fig_same_cov}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{figures/sample_ratio_several_d.eps}
		\caption{Example \ref{ex_sample_ratio}}
		\label{fig_size}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.95\textwidth]{figures/covariate_shift.eps}
		\caption{Example \ref{ex_covshift}}
		\label{fig_covariate}
	\end{subfigure}
	\caption{%Three takeaways of our theory in Section \ref{sec_insight}.
	Validating Example \ref{ex_same_cov} in Section \ref{sec_same} for $10$ tasks:
	In Figure \ref{fig_same_cov}, our estimated loss (solid line) matches the empirical loss (dotted line) accurately for various task-specific variance $d^2$ and sample size $n$ settings. The feature dimension $p$ is $200$, and noise variance $\sigma^2$ is $1/4$.
	Validating Example \ref{ex_sample_ratio} and Example \ref{ex_covshift} in Section \ref{sec_diff} for two tasks:
	in both experiments, our estimated losses (solid line) match the empirical losses (dotted line) accurately.
	In Figure \ref{fig_size}, we discover several interesting phenomena by fixing task two's sample size and increasing task one's sample size.
	Depending on how large $d^2$ is, task two's prediction loss decreases initially before increasing again, or decreases monotonically.
	In Figure \ref{fig_covariate}, we show how different levels of covariate shift affect hard parameter sharing when there is no bias.
	Having covariate shift increases task two's prediction loss when task two's sample size is larger than task one. Otherwise, having covariate shift (surprisingly) decreases task two's prediction loss.}
	\label{fig_model_shift_phasetrans}
\end{figure*}

>>>>

\begin{figure*}[!t]
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/same_covariates.eps}
		\caption{Example \ref{ex_same_cov}}
		\label{fig_same_cov}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/sample_ratio_several_d.eps}
		\caption{Example \ref{ex_sample_ratio}}
		\label{fig_size}
	\end{subfigure}\hfill
	\begin{subfigure}[b]{0.33\textwidth}
		\centering
		\includegraphics[width=0.98\textwidth]{figures/covariate_shift.eps}
		\caption{Example \ref{ex_covshift}}
		\label{fig_covariate}
	\end{subfigure}
	\caption{%Three takeaways of our theory in Section \ref{sec_insight}.
	Our estimated losses (solid line) match the empirical losses (dotted line) accurately under various settings in dimension $p = 200$.
	\textbf{Left.} Validating Example \ref{ex_same_cov} for ten tasks: the noise variance $\sigma^2$ is $1/4$.
	\textbf{Middle.} Validating Example \ref{ex_sample_ratio} for two tasks: we discover several interesting phenomena by fixing task two's sample size and increasing task one's sample size.
	Depending on how large $d^2$ is, task two's prediction loss decreases initially before increasing again, or decreases monotonically.
	\textbf{Right.} We show how different levels of covariate shift affect hard parameter sharing when there is no bias.
	Having covariate shift increases task two's prediction loss when task two's sample size is smaller than task one. Otherwise, having covariate shift (surprisingly) decreases task two's prediction loss.}
	\label{fig_model_shift_phasetrans}
\end{figure*}

### 2020-10-14 arxiv/experiments.tex
typos
various polish

We demonstrate the accuracy of our results via some empirical results.
While our theory is asymptotic (with error terms that are negligible when $p$ is sufficiently large), we observe that they are accurate in a moderate dimension of $p = 200$.

\textbf{Sample efficiency.}
First, we study the setting of Example \ref{ex_same_cov}.
Figure \ref{fig_same_cov} shows the average prediction loss as we increase the number of samples per-task   from $400$ to $2000$.
In all the parameter settings, our results estimate the empirical losses accurately.
We also observe a trend that the average prediction loss increases as we increase $d$ from $0.1$ to $0.2$.
Our work explains the differences between these two settings since $d^2 = 0.1^2$ is always smaller than $\frac{\sigma^2 p}{n - p}$, but $d^2 = 0.2^2$ is not.
%Second, when $d = 0.1$, we have that $d^2 \le \frac{\sigma^2 p}{n - p}$ for all values of $n$, hence the average prediction loss of hard parameter sharing is always lower than STL.
Indeed, we observe a crossover point between hard parameter sharing and STL.
Finally, for $d = 0.2$, looking horizontally at the same prediction loss level, we find that hard parameter sharing requires fewer samples per-task than STL. %\FY{I do not quite understand this sentence about "3x fewer", because how much data needed depends on $d$ and the prediction loss level we are looking at. For example, at the cross point, this ratio is 1. }

\textbf{Sample ratio.}
Second, we show how varying sample ratio impacts hard parameter sharing in the setting of Example \ref{ex_sample_ratio}.
Figure \ref{fig_size} shows task two's prediction loss as we increase the sample ratio $n_1 / n_2$ from $1/10$ to $7/10$.
Again, our estimates are reasonably accurate compared to the empirical losses.
We consider a regime where task two consists of $80,000$ samples, and task one's sample size varies from $8,000$ to $56,000$. 
The task-specific variance is $d^2 = 0.08$, the noise variance is $\sigma^2 = 0.3^2$, and the shared signal variance is $1$. We observe that as we increase the sample ratio, task two's prediction loss decreases initially but later will increases when the sample ratio is above certain level.
On the other hand, when $d^2 = 0.045$, task two's prediction loss decreases monotonically.
Our work explains this trend as discussed below Corollary \ref{cor_MTL_loss}. \HZ{check}

\textbf{Covariate shift.}
Finally, we show how varying covariate shift impacts hard parameter sharing in the setting of Example \ref{ex_covshift}. 
Figure \ref{fig_covariate} shows task two's prediction loss as we increase task one's sample size.
Recall that $\lambda$ measures the severity of covariate shifts---a larger $\lambda$ means a larger covariate shift.
Our estimates match the empirical losses accurately, and indeed we observe the dichotomy in Example \ref{ex_covshift}.
The noise variance $\sigma^2$ is $1/4$.
%Covariate shift: We set $\kappa = 1$ and $d = 0$.
%We set $\rho_2 = 4$ and vary $\rho_1$ from $5$ to $25$ for sample sizes.
%We use the scale parameter $\lambda = 1$ for the curve without covariate shift and $\lambda = 2$ for the curve with covariate shift (cf. Section \ref{sec_covshift}).

>>>>

We demonstrate the accuracy of our results in simulations.
While our theory is asymptotic (with error terms that are negligible when $p$ is sufficiently large), we observe that they are incredibly accurate in a moderate dimension of $p = 200$.

\textbf{Sample efficiency.}
First, we study the setting of Example \ref{ex_same_cov}.
Figure \ref{fig_same_cov} shows the average prediction loss as we increase the number of samples per-task   from $400$ to $2000$.
In all the parameter settings, our results estimate the empirical losses accurately.
We also observe a trend that the average prediction loss increases as we increase $d$ from $0.1$ to $0.2$.
Our work explains the differences between these two settings since $d^2 = 0.1^2$ is always smaller than $\frac{\sigma^2 p}{n - p}$, but $d^2 = 0.2^2$ is not.
%Second, when $d = 0.1$, we have that $d^2 \le \frac{\sigma^2 p}{n - p}$ for all values of $n$, hence the average prediction loss of hard parameter sharing is always lower than STL.
Indeed, we observe a crossover point between hard parameter sharing and STL.
Finally, for $d = 0.2$, looking horizontally, we find that HPS requires fewer samples per-task than STL to achieve the same loss level. %\FY{I do not quite understand this sentence about "3x fewer", because how much data needed depends on $d$ and the prediction loss level we are looking at. For example, at the cross point, this ratio is 1. }

\textbf{Sample ratio.}
Second, we show how varying sample ratio impacts hard parameter sharing in the setting of Example \ref{ex_sample_ratio}.
Figure \ref{fig_size} shows task two's prediction loss as we increase the sample ratio $n_1 / n_2$ from $1/10$ to $7/10$.
%Again, our estimates are accurate compared to the empirical losses.
We consider a regime where task two consists of $80,000$ samples, and task one's sample size varies from $8,000$ to $56,000$. 
The task-specific variance is $d^2 = 0.08$, the noise variance is $\sigma^2 = 0.3^2$, and the shared signal variance is $1$. We observe that as we increase the sample ratio, task two's prediction loss decreases initially but later will increase when the sample ratio is above a certain level.
On the other hand, when $d^2 = 0.045$, task two's prediction loss decreases monotonically.
Our result from Corollary \ref{cor_MTL_loss} explains this trend. \HZ{check}

\textbf{Covariate shift.}
Finally, we show how varying covariate shift impacts hard parameter sharing in the setting of Example \ref{ex_covshift}. 
Figure \ref{fig_covariate} shows task two's prediction loss as we increase task one's sample size.
Recall that $\lambda$ measures the severity of covariate shifts---a larger $\lambda$ means a larger covariate shift.
We indeed observe the dichotomy in Example \ref{ex_covshift} at $n_1 = 800$.
The sample size $n_2$ is $800$ and the noise variance $\sigma^2$ is $1/4$.
%Covariate shift: We set $\kappa = 1$ and $d = 0$.
%We set $\rho_2 = 4$ and vary $\rho_1$ from $5$ to $25$ for sample sizes.
%We use the scale parameter $\lambda = 1$ for the curve without covariate shift and $\lambda = 2$ for the curve with covariate shift (cf. Section \ref{sec_covshift}).

### 2020-10-14 arxiv/setup.tex
various polish

%We assume that there are multiple datasets that all follow (potentially different) linear models.
%In each dataset, suppose the feature vector is $x = {\Sigma}^{1/2} z$, where $z \in \real^p$ has i.i.d entries with zero mean and unit variance, and the population covariance matrix $\Sigma \in\real^{p \times p}$ is deterministic and positive semidefinite.
Suppose we have $t$ datasets.
For each dataset $i$ from $1$ to $t$, let $n_i$ denote its sample size.
Let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates.
We assume that the label vector $Y^{(i)} \in \real^{n_i}$ follows a linear model plus random noise.
We study the standard hard parameter sharing architecture to learn jointly from multiple datasets.
In this architecture, there is a shared feature representation layer $B\in\real^{p\times r}$ for all tasks and a separate output layer $A_i \in \real^r$ for every task $i$.
See Figure \ref{fig_intro_arch} for an illustration.
%The width of $B$, denoted by $r$, plays an important role in regularization.
%As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
%Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
We study the following minimization problem:
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
%	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
%\vspace{-0.1in}
\begin{align}\label{eq_mtl}
			f(A, B) = \sum_{i=1}^t \norm{X^{(i)} B A_i - Y^{(i)}}^2,
\end{align}
%\vspace{-0.1in}
where $A = [A_1, A_2, \dots, A_t] \in \real^{r \times t}$.
Given a solution $(\hat{A}, \hat{B})$ of $f(A, B)$ (which we will specify later), let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the HPS estimator for task $i$.
The critical questions here are:
(i) How well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) For datasets with different sample sizes and covariate shifts, how do they affect the estimator?

\textbf{Main results.}
Our first result (Theorem \ref{thm_many_tasks}) applies to the multi-label prediction setting where all datasets have the same features, and we want to make multiple predictions (see e.g. \cite{hsu2009multi}).
We analyze the global minimizer of $f(A, B)$, and provide a sharp generalization bound on its (out-of-sample) prediction loss for any task.
This case is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
We show that the prediction loss of HPS admits a clean bias-variance decomposition.
Our results imply that in the same covariates setting where the sample sizes are also equal, HPS helps by reducing the variance compared to single-task learning, but hurts by increasing the bias.

@@

\textit{Covariate shift (Example \ref{ex_covshift})}: In addition to sample sizes, variance also scales with the covariate shift between different datasets. For a large sample size ratio, HPS's  variance is smallest when there is no covariate shift. Counterintuitively, for a small sample size ratio, having covariate shifts reduces variance through a complementary spectrum.
%\end{enumerate}

@@

\textbf{Organizations.}
The rest of this paper is organized as follows.
In Section \ref{sec_same}, we present the bias-variance decomposition for hard parameter sharing.
In Section \ref{sec_diff}, we present our technical results that describe how varying sample sizes and covariate shifts impact hard parameter sharing using random matrix theory.
In Sections \ref{sec_simulation} and \ref{sec_text}, we validate our theory in both simulations and a real world classification task.
In Section \ref{sec_conclude}, we conclude the paper and describe several open questions.

\textbf{Notations.}
%Let $\cE \define [\varepsilon_1, \varepsilon_2, \dots, \varepsilon_t] \in \real^{n \times t}$ denote the random noise.
%We can also write $Y = XB^{\star} + \cE$.
%Let $A = [A_1, A_2, \dots, A_t] \in \real^{r\times t}$ be a matrix notation that contains all the output layer parameters.
For an $n\times p$ matrix $X$, let $\lambda_{\min}(X)$ denote its smallest singular value and $\norm{X}$ denote its spectral norm.
Let $\lambda_1(X) \ge \lambda_2(X) \ge \cdots \ge \lambda_{p\wedge n}(X)$ denote the singular values of $X$.
Let $X^+$ denote the Moore-Penrose psuedoinverse.
We refer to random matrices of the form $\frac {X^\top X} n$ as sample covariance matrices.
We say that an event $\Xi$ holds with high probability if the probability that $\Xi$ happens goes to $1$ as $p$ goes to infinity.
%We shall use $\oo(1)$ to mean a small positive quantity that converges to 0 as $p$ goes to infinity.

>>>>

%We assume that there are multiple datasets that all follow (potentially different) linear models.
%In each dataset, suppose the feature vector is $x = {\Sigma}^{1/2} z$, where $z \in \real^p$ has i.i.d entries with zero mean and unit variance, and the population covariance matrix $\Sigma \in\real^{p \times p}$ is deterministic and positive semidefinite.
Suppose we have $t$ datasets.
For each dataset $i$ from $1$ to $t$, let $n_i$ denote its sample size.
Let $X^{(i)} \in \real^{n_i \times p}$ denote dataset $i$'s feature covariates.
We assume that the label vector $Y^{(i)} \in \real^{n_i}$ follows a linear model plus random noise.
We study the standard hard parameter sharing architecture:
there is a shared feature representation layer $B\in\real^{p\times r}$ for all tasks and a separate output layer $A_i \in \real^r$ for every task $i$.
See Figure \ref{fig_intro_arch} for an illustration.
%The width of $B$, denoted by $r$, plays an important role in regularization.
%As observed in Proposition 1 of \citet{WZR20}, if $r \ge t$, there is no regularization effect.
%Hence, we assume that $r < t$ in our study.
%For example, when there are only two tasks, $r = 1$ and $B$ reduces to a vector whereas $W_1, W_2$ become scalars.
We study the following minimization problem:
%	\item Separate each dataset $(X_i, Y_i)$ randomly into a training set $(X_i^{tr}, Y_i^{tr})$ and a validation set $(X_i^{val}, Y_i^{val})$.
%	The size of each set is described below.
%	\item Learn the shared layer $B$: minimize the training loss over $B$ and $W_1, \dots, W_t$, leading to a local minimum of $B$ that depends on $W_1, \dots, W_t$, denoted by $\hat{B} = \hat{B}(W_1, \dots, W_t)$.
%\vspace{-0.1in}
\begin{align}\label{eq_mtl}
			f(A, B) = \sum_{i=1}^t \norm{X^{(i)} B A_i - Y^{(i)}}^2,
\end{align}
%\vspace{-0.05in}
where $A = [A_1, A_2, \dots, A_t] \in \real^{r \times t}$.
Given a solution $(\hat{A}, \hat{B})$ of $f(A, B)$ (which we will specify later), let $\hat{\beta}_i^{\MTL} = \hat{B} \hat{A}_i$ denote the HPS estimator for task $i$.
The critical questions here are:
(i) How well does the estimator work? In particular, how does the performance of the estimator scale with sample size?
(ii) For datasets with different sample sizes and covariate shifts, how do they affect the estimator?

\textbf{Main results.}
Our first result (Theorem \ref{thm_many_tasks}) applies to multi-label prediction settings where all datasets have the same features (and sample size), and we want to make multiple predictions (see e.g. \cite{hsu2009multi}).
We analyze the global minimizer of $f(A, B)$, and provide a sharp bias-variance decomposition of its (out-of-sample) prediction loss for any task.
This setting is tractable even though in general, $f(A, B)$ is non-convex in $A$ and $B$ (e.g. matrix completion is a special case for suitably designed $X^{(i)}, Y^{(i)}$).
%We show that the prediction loss of HPS admits a clean bias-variance decomposition.
Our result implies that when all tasks have the same covariates, HPS helps by reducing variance compared to single-task learning, but hurts by increasing bias.

@@

\textit{Covariate shift (Example \ref{ex_covshift})}: In addition to sample sizes, variance also scales with two datasets' covariate shifts. For a large sample size ratio, HPS's  variance is smallest when there is no covariate shift. Counterintuitively, for a small sample size ratio, having covariate shifts reduces variance through a complementary spectrum. We achieve this result through a novel characterization on the sum of two sample covariance matrices with arbitrary covariate shifts. See our discussion of proof techniques below for more details.
%\end{enumerate}

@@

\textbf{Organizations.}
The rest of this paper is organized as follows.
In Section \ref{sec_same}, we present the bias-variance decomposition for hard parameter sharing.
In Section \ref{sec_diff}, we present our technical results that describe how varying sample sizes and covariate shifts impact hard parameter sharing using random matrix theory.
In Sections \ref{sec_simulation} and \ref{sec_text}, we validate our theory in simulations and real world classification tasks.
In Section \ref{sec_conclude}, we conclude the paper and describe several open questions.
Section \ref{app_proof_error_same_cov} and \ref{appendix RMT} present proofs of our results.

\textbf{Notations.}
%Let $\cE \define [\varepsilon_1, \varepsilon_2, \dots, \varepsilon_t] \in \real^{n \times t}$ denote the random noise.
%We can also write $Y = XB^{\star} + \cE$.
%Let $A = [A_1, A_2, \dots, A_t] \in \real^{r\times t}$ be a matrix notation that contains all the output layer parameters.
For an $n\times p$ matrix $X$, let $\lambda_{\min}(X)$ denote its smallest singular value and $\norm{X}$ denote its largest singular value.
Let $\lambda_1(X), \lambda_2(X), \cdots, \lambda_{p\wedge n}(X)$ denote the singular values of $X$ in decreasing order.
Let $X^+$ denote the Moore-Penrose psuedoinverse of $X$.
We refer to random matrices of the form $\frac {X^\top X} n$ as sample covariance matrices.
We say that an event $\Xi$ holds with high probability if the probability that $\Xi$ happens goes to $1$ as $p$ goes to infinity.
%We shall use $\oo(1)$ to mean a small positive quantity that converges to 0 as $p$ goes to infinity.
